Sensitivity Analysis

### 2024-09-18 Article_Sensitivity_Analsysis-Chamomile.tex
18.09

@@

Sensitivity analysis is a tool used to understand how model parameters affect its output. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. This work implemented the automatic differentiation to derive the sensitivity equations. The local sensitivity analysis techniques consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions. In the case of dynamical systems, local sensitivity analysis provides a time series describing how that dependency evolves.

The local sensitivity analysis can be performed with respect to any model parameter, but this work focuses on the effect of pressure increase. At selected operating conditions ($35^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s), the pressure increase enhances the mass transfer, which leads to faster loss of solute from particles and consequently to negative sensitivities in the solid phase. Analogously, the sensitivities in the fluid phase are characterised by positive deviations, which indicates that more solute is transported to the fluid phase. As a result, the extraction yield is improved and characterised by positive sensitivities. As the results of the local sensitivity analysis depends on the operating point, the analysis was repeated at different nominal pressures. It can be observed that system responses at low-pressures are stronger than at high-pressures. This behaviour can be explained by more rapid changes of the solvent properties close to the critical point and lower process efficiency at low-pressure.

Local sensitivity analysis can provide valuable information by identifying which parameters are influential and how these influence changes over time. Sensitivity analysis can highlight discrepancies between model predictions and actual experimental results. Identifying unexpected relationships or non-linearities can prompt further investigation and refinement of the process model. Sensitivity analysis can inform the design of future experiments by indicating which parameters should be varied and monitored.

% ===================================================
% Bibliography
% ===================================================
%% Loading bibliography style file
%\clearpage
%\bibliographystyle{model1-num-names}
\bibliographystyle{unsrtnat}
\bibliography{mybibfile}

@@

>>>>

\usepackage{nomencl}
\makenomenclature

\usepackage{etoolbox}
\renewcommand\nomgroup[1]{%
	\item[\bfseries
	\ifstrequal{#1}{A}{Latin symbols}{%
		\ifstrequal{#1}{B}{Greek symbols}{%
			\ifstrequal{#1}{C}{Abberivations}{
	}}}%
	]}

@@

Sensitivity analysis is a tool used to understand how model parameters affect its output. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. This work implemented the automatic differentiation to derive the sensitivity equations. The local sensitivity analysis techniques consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions. In the case of dynamical systems, the outcome of local sensitivity analysis quantifies how the state of the system or its output varies in time with respect to small perturbations in parameters.

The local sensitivity analysis can be performed with respect to any model parameter, but this work focuses on the effect of pressure increase. At selected operating conditions ($35^\circ C$ and $5\cdot 10^{-5}$ kg/s), the pressure increase enhances the mass transfer, which leads to faster loss of solute from particles and consequently to negative sensitivities in the solid phase. Analogously, the sensitivities in the fluid phase are characterised by positive deviations, which indicates that more solute is transported to the fluid phase. As a result, the extraction yield is improved and characterised by positive sensitivities. As the results of the local sensitivity analysis depends on the operating point, the analysis was repeated at different nominal pressures. It can be observed that system responses at low-pressures are stronger than at high-pressures. This behaviour can be explained by rapid changes of the solvent properties close to the critical point.

Local sensitivity analysis results provide valuable information by identifying which parameters are influential and how these influence changes over time. One application of the sensitivity analysis is to spot discrepancies between model predictions and actual experimental results. Identifying unexpected system responses prompts further investigation and refinement of the process model. Sensitivity analysis results can be used to design future experiments by indicating which parameters should be varied and monitored. Moreover, the parameters which are characterized by low sensitivity can be subject of model reduction.

% ===================================================
% Bibliography
% ===================================================
%% Loading bibliography style file
%\clearpage
\newpage
%\bibliographystyle{model1-num-names}
\bibliographystyle{unsrtnat}
\bibliography{mybibfile}

@@

\newpage

\nomenclature[A]{\(P\)}{Pressure}
\nomenclature[A]{\(T\)}{Temperature}
\nomenclature[A]{\(T^{in}\)}{Inlet temperature}
\nomenclature[A]{\(T^{out}\)}{Outlet temperature}
\nomenclature[A]{\(F\)}{Mass flow rate}
\nomenclature[A]{\(u\)}{Superfical velocity}
\nomenclature[A]{\(v\)}{Linear velocity}
\nomenclature[A]{\(z\)}{Spatial direction}
\nomenclature[A]{\(e\)}{Internal energy}
\nomenclature[A]{\(h\)}{Enthalpy}
\nomenclature[A]{\(t\)}{Time}
\nomenclature[A]{\(t_f\)}{Total extraction time}
\nomenclature[A]{\(t_0\)}{Inital extraction time}
\nomenclature[A]{\(A\)}{Total cross section of the bed}
\nomenclature[A]{\(A_f\)}{Cross section of the bed accupied by the fluid}
\nomenclature[A]{\(c_f\)}{Concentration of solute in fluid phase}
\nomenclature[A]{\(c_{f0}\)}{Inital concentration of solute in fluid phase}
\nomenclature[A]{\(c_f^*\)}{Concentration of solute at the solid-fluid interface}
\nomenclature[A]{\(c_s\)}{Concentration of solute in solid phase}
\nomenclature[A]{\(c_{s0}\)}{Inital concentration of solute in solid phase}
\nomenclature[A]{\(c_s^*\)}{Concentration of solute at the solid-fluid interface}
\nomenclature[A]{\(c_p\)}{Concentration of solute in the core of a pore}
\nomenclature[A]{\(c_{pf}\)}{Concentration of solute in the pore opening}
\nomenclature[A]{\(D_e^M\)}{Axial diffusion coefficeint}
\nomenclature[A]{\(r_e\)}{Mass transfer kinetic term}
\nomenclature[A]{\(L\)}{Length of fixed bed}
\nomenclature[A]{\(S\)}{Sensitivity equations}
\nomenclature[A]{\(\textbf{G}\)}{Augumented system}
\nomenclature[A]{\(\dot{S}\)}{Time derivative of sensitivity equations}
\nomenclature[A]{\(\bar{S}\)}{Sensitivity matrix}
\nomenclature[A]{\(\bar{J}_x\)}{Jacobian matrix with respect to the state space}
\nomenclature[A]{\(\bar{J}_\Theta\)}{Jacobian matrix with respect to the paramters}
%\nomenclature[A]{\(\mathcal{L}\)}{Log-likelihood function}
\nomenclature[A]{\(l\)}{Characterisitic diemsinon of particles}
\nomenclature[A]{\(D_i\)}{Interanl diffusion coefficient}
\nomenclature[A]{\(D_i^R\)}{Reference value of interanl diffusion coefficient}
\nomenclature[A]{\(r\)}{Particle radius}
\nomenclature[A]{\(k_p\)}{Volumetric partition coefficient}
\nomenclature[A]{\(k_m\)}{Mass partition coefficient}
\nomenclature[A]{\(y\)}{Extraction yield}
\nomenclature[A]{\(Y\)}{Yield measurment}
\nomenclature[A]{\(x\)}{State vector}
\nomenclature[A]{\(G\)}{Vector of discretized differential equations}
\nomenclature[A]{\(p\)}{Probaility disribution model}
\nomenclature[A]{\(R_e\)}{Reynolds number}

\nomenclature[B]{\(\rho_f\)}{Fluid density}
\nomenclature[B]{\(\Phi\)}{Bed porosity}
\nomenclature[B]{\(\rho_s\)}{Bulk density of solid}
\nomenclature[B]{\(\mu\)}{Sphericity coefficient}
\nomenclature[B]{\(\gamma\)}{Decaying function}
\nomenclature[B]{\(\Upsilon\)}{Decay coefficient}
\nomenclature[B]{\(\Theta\)}{Paramter space}
\nomenclature[B]{\(\theta\)}{vector of unknown parameters}
\nomenclature[B]{\(\epsilon\)}{Unobervable error}
\nomenclature[B]{\(\sigma\)}{Standard deviation}

\nomenclature[C]{BIC}{Broke-and-Intact Cell model}
%\nomenclature[C]{MLE}{Maximum Likelihood Estimation}
\nomenclature[C]{SFE}{Supercritical Fluid Extraction}
\nomenclature[C]{HBD}{Hot Ball Diffusion}
\nomenclature[C]{SC}{Shrinking Core}

\printnomenclature

### 2024-09-18 Sections/Sensitivity_Pressure.tex
18.09

	As discussed in Chapter \ref{CH:Governing_equations_chapter}, a small pressure wave propagates at the speed of sound relative to the flow. When the flow velocity is relatively low, pressure changes are hydrodynamic (resulting from velocity motion) rather than thermodynamic. The Low Mach number assumption allows for the instant propagation of thermodynamic pressure throughout the system, enabling a single pressure value to be considered for the entire system. In response to a pressure change, the energy equation experiences a simultaneous deviation across the entire spatial domain. This pressure change impacts the fluid's temperature within the computational domain, while boundary values are constrained by conditions specified at the domain's extremes. Dirichlet boundary conditions impose a fixed temperature value at the inlet, creating a thermal gradient that propagates through the system. Conversely, Neumann boundary conditions specify the heat flux at the boundaries. In this study, zero Neumann boundary conditions were applied to ensure that the temperatures at the inlet, outlet, and within the extractor vary uniformly in response to the pressure change.

@@

	Figure \ref{fig:Sensitivty_P_y} show how sensitive the extraction yield is to pressure changes over an extended period for various pressure values. Initially, all sensitivity curves remain nearly flat, indicating a delayed response in the system. Due to the reduced fluid velocity, the solute reaches the extractor's outlet later, causing minor negative sensitivities to appear. Once the solute exits the extractor, the sensitivity curves rise rapidly. Positive yield sensitivities indicate improved process efficiency and enhanced mass transfer. The peak in $\frac{dy}{dP_{in}}$ represents the point of greatest deviation from the original system. Eventually, the sensitivities decline and converge towards zero as the concentration gradient becomes a limiting factor, reducing the impact of enhanced mass transfer.

	As presented in Figure \ref{fig:Sensitivty_P_y}, higher sensitivities are observed for the low-pressure system. At lower pressures, the supercritical fluid has lower density and solvating power, which results in less efficient extraction as shown by data given in {\color{red}article 1}. Small changes in pressure at low pressures can significantly impact the solute's solubility and, consequently, the extraction yield. Moreover, near the critical point, small changes in pressure can lead to significant changes in the physical properties of the supercritical fluid, such as density and viscosity. These changes can strongly affect the solute's solubility and the mass transfer rates, resulting in higher sensitivity at lower pressures where the system operates closer to the critical point.

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/Yield_Multiple.png}
		\caption{The effect of $P_{in}$ change on $y(t)$}
		\label{fig:Sensitivty_P_y}
	\end{figure}

>>>>

	As discussed in Chapter \ref{CH:Governing_equations_chapter}, a small pressure wave propagates at the speed of sound relative to the flow. When the flow velocity is relatively low, pressure changes are hydrodynamic (resulting from velocity motion) rather than thermodynamic. The Low Mach number assumption allows for the instant propagation of thermodynamic pressure throughout the system, enabling a single pressure value to be considered for the entire system. In response to a pressure change, the energy equation experiences a simultaneous deviation across the entire spatial domain. This pressure change impacts the fluid's temperature within the computational domain, while boundary values are constrained by conditions specified at the domain's extremes. Dirichlet boundary conditions impose a fixed temperature value at the inlet, creating a thermal gradient that propagates through the system. On the other hand, Neumann boundary conditions specify the heat flux at the boundaries. In this study, zero Neumann boundary conditions were applied to ensure that the temperatures at the inlet, outlet, and within the extractor vary uniformly in response to the pressure change.

@@

	Figure \ref{fig:Sensitivty_P_y} show how sensitive the extraction yield is to pressure changes over an extended period for various pressure values. Initially, all sensitivity curves remain nearly flat, indicating a delayed response in the system. Due to the reduced fluid velocity, the solute reaches the extractor's outlet later, causing minor negative sensitivities to appear. Once the solute exits the extractor, the sensitivity curves rise. Positive yield sensitivities indicate improved process efficiency and enhanced mass transfer. The peak in $\frac{dy}{dP_{in}}$ represents the point of greatest deviation from the original system. Eventually, the sensitivities decline and converge towards zero as the concentration gradient becomes a limiting factor, reducing the impact of enhanced mass transfer.

	As presented in Figure \ref{fig:Sensitivty_P_y}, higher sensitivities are observed for the low-pressure system. At lower pressures, the supercritical fluid has lower density and solvating power, which results in less efficient extraction as shown by data given in {\color{red}article 1}. Small changes in pressure at low pressures can significantly impact the solute's solubility and, consequently, the extraction yield. Moreover, near the critical point, small changes in pressure can lead to significant changes in the physical properties of the supercritical fluid, such as density and viscosity and consequently to higher sensitivity of the system state-space.

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.1cm 0.0cm 1.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/Yield_Multiple.png}
		\caption{The effect of $P_{in}$ change on $y(t)$}
		\label{fig:Sensitivty_P_y}
	\end{figure}

### 2024-09-18 Sections/introduction_imp.tex
Updates from Overleaf
18.09

	This study investigates the extraction of essential oil from chamomile flowers (Matricaria chamomilla L.) via supercritical fluid extraction techniques and the modelling of this process. Chamomile is a medicinal herb widely cultivated in southern and eastern Europe—such as Germany, Hungary, France, and Russia. It can be found outside of Europe in Brazil as discussed by \citet{Singh2011}. This plant is distinguished by its hollow, bright gold cones, housing disc or tubular florets and surrounded by about fifteen white ray or ligulate florets. Chamomile has been used for its medicinal benefits, serving as an anti-inflammatory, antioxidant, mild astringent, and healing remedy. Chamomile's aqueous extract is widely used to calm nerves and mitigate anxiety, hysteria, nightmares, insomnia, and other sleep-related conditions, according to \citet{Srivastava2009}. \citet{Orav2010} reported that oil yields from dried chamomile samples ranged from 0.7 to 6.7 mL/kg. The highest yields of essential oil, between 6.1 and 6.7 mL/kg, were derived from chamomile sourced from Latvia and Ukraine. In comparison, chamomile from Armenia exhibited a lower oil content of 0.7 mL/kg.

	Evaluating the economic viability of the process is essential when choosing the suitable technology for essential oil extraction. Traditional methods, such as distillation and organic solvent extraction, are commonly employed but have drawbacks. Distillation, for example, involves high temperatures that can lead to the thermal degradation of heat-sensitive compounds. This limitation has led to the increased popularity of alternative techniques like supercritical fluid extraction. Supercritical carbon dioxide is appealing due to its distinctive properties: it is inflammable, non-toxic, and non-corrosive. Supercritical fluids can exhibit both gas- and liquid-like properties, allowing for adjustable dissolving power through changes in operating conditions.

	The literature offers various mathematical models to describe the extraction of valuable compounds from biomass. Selecting a process model is case-to-case dependent and requires analysing each model's specific assumptions about mass transfer and thermodynamic equilibrium.

	The model proposed by \citet{Reverchon1993} is called the hot ball model, as it is based on an analogy to heat transfer and describes an extraction process from solid particles. This model assumes that particles contain low quantities of solute and that solubility is not a limiting factor.

	The Broken-and-Intact Cell model, proposed by \citet{Sovova1994}, assumes that external surfaces of particles are mechanically disrupted, allowing the solvent's access to the solute in broken cells. In contrast, the solute in intact cells remains less accessible due to higher mass transfer resistance.

	\citet{Reverchon1996} formulated a fluid-solid extraction model where the solute is treated as a single component, governed by internal mass transfer resistance and omitting the effects of external mass transfer, axial dispersion, and variations in fluid density and flow rate throughout the bed.

	This work builds upon the linear kinetic model suggested by \citet{Reverchon1996}, deriving fundamental governing equations to develop a comprehensive model for the chamomile oil extraction process. This model aims for control-oriented simplicity, assuming a semi-continuous operation within a cylindrical vessel. The process involves supercritical solvent being pumped through a fixed bed of finely chopped biomass to extract the solute, followed by separation of the solvent and solute in a flush drum to collect the extract. Parameters such as the pressure ($P$), feed flow rate ($F_{in}$) and inlet temperature ($T_{in}$) are adjustable and measurable, while the outlet temperature ($T_{out}$) and the amount of product at the outlet can only be monitored. Figure \ref{fig: SFE_drawing} presents a simplified process flow diagram.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/PFD.drawio.pdf}
		\caption{Process flow diagram}
		\label{fig: SFE_drawing}
	\end{figure}

>>>>

	Supercritical CO$_2$ is defined as carbon dioxide that is pressurized and heated above its critical point (31.1 $^\circ C$, 74 bar). Depending on the operating conditions, the fluid properties such as viscosity and density can vary, which leads to multiple industrial applications of CO$_2$.

	The supercritical carbon dioxide is commonly used for impregnation as described by \citet{Weidner2018}, \citet{Machado2022} or \citet{Fathi2022}. Impregnation is defined as modifying the properties of bulk substances by physically or chemically binding/adsorbing impregnates to a bulk material or surface, such as the hydrophobization of surfaces. The main advantage of using supercritical CO$_2$ is that after depressurization, it desorbs from the surface and evaporates, leaving a solvent-free product. On the other hand, the main disadvantage of using carbon dioxide for impregnation is the low solubility of many drugs of interest.

	Another application of supercritical CO$_2$ is nanoparticles formation as investigated by \citet{Padrela2018}, \citet{Franco2021}, \citet{SaadatiArdestani2020} or \citet{Sodeifian2022}. Supercritical carbon-dioxide-assisted technologies enable the production of different morphologies of different sizes, including nanoparticles and nanocrystals, by modulating operating conditions. Supercritical fluid-based processes have advantages over techniques conventionally employed to produce nanosized particles or crystals, such as reduced use of toxic solvents. Moreover, the CO$_2$ is completely removed from the final product by simple depressurization.

	One of the most popular applications of supercritical CO$_2$ is the extraction of essential oils, as described by many researchers, for example, by \citet{Sodeifian2017}, \citet{Reverchon1993} or \citet{Sovova1994}. Traditional methods, such as distillation and organic solvent extraction, are commonly employed but have drawbacks. Distillation, involves high temperatures that can lead to the thermal degradation of heat-sensitive compounds. This limitation has increased the popularity of alternative techniques, such as supercritical fluid extraction. Supercritical CO$_2$ is appealing due to its distinctive properties: it is inflammable, non-toxic and non-corrosive. Supercritical fluids can exhibit both gas- and liquid-like properties, allowing for adjustable dissolving power through changes in operating conditions.

	This study investigates the extraction of essential oil from chamomile flowers (Matricaria chamomilla L.) via supercritical fluid extraction techniques and the modelling of this process. Chamomile is a medicinal herb widely cultivated in southern and eastern Europe — in countries such as Germany, Hungary, France and Russia. It can be found outside Europe, for instance in Brazil as discussed by \citet{Singh2011}. This plant is distinguished by its hollow, bright gold cones, housing disc or tubular florets and surrounded by about fifteen white ray or ligulate florets. Chamomile has been used for its medicinal benefits, serving as an anti-inflammatory, antioxidant, mild astringent, and healing remedy. Extracts of chamomile are widely used to calm nerves and mitigate anxiety, hysteria, nightmares, insomnia and other sleep-related conditions, according to \citet{Srivastava2009}. \citet{Orav2010} reported that oil yields from dried chamomile samples ranged from 0.7 to 6.7 mL/kg. The highest yields of essential oil, between 6.1 and 6.7 mL/kg, were derived from chamomile sourced from Latvia and Ukraine. In comparison, chamomile from Armenia exhibited a lower oil content of 0.7 mL/kg.

	The literature offers various mathematical models to describe the extraction of valuable compounds from biomass. Selecting a process model is case-to-case dependent and requires analysis of each model's specific assumptions about mass transfer and thermodynamic equilibrium.

	\citet{Goto1996} presented the Shrinking Core (SC) model, which describes a process of irreversible desorption that is followed by diffusion through the pores of a porous solid. When the mass transfer rate of the solute in the non-extracted inner region is significantly slower than in the outer region, where most of the solute has already been extracted, or when the solute concentration exceeds its solubility in the solvent, a distinct boundary may form between the inner and outer regions. As extraction progresses, the core of the inner region shrinks. The model envisions supercritical CO$_2$ extraction as a sharp, inward-moving front, with a completely non-extracted core ahead of the front and a fully extracted shell behind it.

	\citet{Sovova1994} proposed The Broken-and-Intact Cell (BIC) model, which assumes that a portion of the solute, initially stored within plant structures and protected by cell walls, is released during the mechanical breakdown of the material. The solute located in the region of broken cells near the particle surface is directly exposed to the solvent, while the core of the particle contains intact cells with undamaged walls. This model describes three extraction phases: a fast extraction phase for accessible oil, a transient phase, and a slow phase controlled by diffusion. The model has been successfully applied to the extraction of grape oil (\citet{Sovova1994b}) and caraway oil (\citet{Sovova1994a}).

	The Supercritical Fluid Extraction (SFE) process can be be treated similarly to heat transfer, considering solid particles like hot balls cooling down in a uniform environment. \citet{Bartle1990} introduced the hot ball diffusion (HBD) model, where spherical particles with uniformly distributed solute diffuse similarly to heat diffusion. Unlike the BIC model, where solute is readily available on the particle surface, the HBD model is suited for systems with small quantities of extractable materials and is not limited by solubility. The model is particularly relevant when internal diffusion controls mass transfer, allowing results from single particles to be extended to the entire bed under uniform conditions. \citet{Reverchon1993} have further elaborated on the HBD model and used it to simulate extraction processes for natural materials.

	\citet{Reverchon1996} proposed a model for extraction of essential oils, which are mainly located inside the vegetable cells in organules called vacuoles. Only a small fraction of essential oil might be near the particle surface due to the breaking up of cells during grinding or in epidermal hairs located on the leaf surface. The fraction of oil freely available on the particle surface should not be significant in the case of SFE from leaves. Consequently, SFE of essential oil from leaves should be mainly controlled by the internal mass-transfer resistance. Therefore, the external mass-transfer coefficient was neglected in the development of the model of \citet{Reverchon1996}. The mass balances were developed in the additional hypotheses that the axial dispersion can be neglected and that the solvent density and flow rate are constant along the bed.

	This work builds upon the linear kinetic model suggested by \citet{Reverchon1996}, deriving fundamental governing equations to develop a comprehensive model for the chamomile oil extraction process. This model aims for control-oriented simplicity, assuming a semi-continuous operation within a cylindrical vessel. The process involves a supercritical solvent being pumped through a fixed bed of finely chopped biomass to extract the solute, followed by separation of the solvent and solute in a flush drum to collect the extract. Parameters such as pressure ($P$), feed flow rate ($F$) and inlet temperature ($T^{in}$) are adjustable and measurable, while the outlet temperature ($T^{out}$) and the amount of product at the outlet can only be monitored. Figure \ref{fig: SFE_drawing} presents a simplified process flow diagram.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/PFD.drawio.pdf}
		\caption{Process flow diagram.}
		\label{fig: SFE_drawing}
	\end{figure}
